Power to the Stimuli: Not the Effect

### 2023-03-23 manuscript_draft.Rmd
bam bam bam

abstract: |
  We will add an abstract that explains that we are going to 1) talk about this cool procedure, 2) show some simulations that the procedure works, and 3) give two examples of the procedure in action. You should try it out!

@@

# bibliography      : ["r-references.bib"]

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_docx
---

@@

This part we will put in a real introduction. Things we need to discuss:

  - Repeated measures data with items
  - Accuracy in Parameter Estimation 
  - Adaptive sampling
  - MLM designs account for item heterogeneity but a lot of people anova

@@

Using ideas from accuracy in parameter estimation, we suggest the following procedure to determine a sample size for each item:

@@

To explore if we could recover the new projected sample size from data a researcher would have, we used linear models to create a formula for researcher correction. First, the corrected projected sample size was predicted by the original projected sample size. Next, the standard deviation of the item standard deviations was added to the equation to recreate heterogeneity estimates. The scale of the data is embedded into the standard deviation of the items (*r* = `r round(cor(summary_long$scale, summary_long$sd_item), digits = 2)`), and therefore, this variable was not included separately. Last, we included the pilot sample size. 
```{r regressions, include = F, echo = F}
user_model <- lm(new_sample ~ sample_size, data = summary_long)
user_print <- apa_print(user_model)
user_sum <- summary(user_model)

@@

The first model using pilot sample size to predict new sample size was significant, *F*(`r user_sum$fstatistic[2]`, `r user_sum$fstatistic[3]`) = `r round(user_sum$fstatistic[1], digits = 2)`, *p* < .001, `r user_print$estimate$modelfit$r2`, capturing nearly 90% of the variance, `r user_print$estimate$sample_size`. The second model with item standard deviation was better than the first model *F*(`r change_table$df[2]`, `r format(change_table$res.df[2], big.mark = "")`) = `r round(change_table$statistic[2], digits = 2)`, *p* < .001, `r user_print2$estimate$modelfit$r2`. The item standard deviation predictor was significant, `r user_print2$full_result$sd_item`. The addition of the original pilot sample size was also significant, *F*(`r change_table$df[3]`, `r format(change_table$res.df[3], big.mark = "")`) = `r round(change_table$statistic[3], digits = 2)`, *p* < .001, `r user_print3$estimate$modelfit$r2`.

As shown in the final model Table \@ref(tab:table-correction), the new suggested sample size is proportional to the original suggested sample size (i.e., *b* < 1), which reduces the sample size suggestion. As variability increases, the suggested sample size also increases to capture differences in heterogeneity shown above; however, this predictor is not significant in the final model, and only contributes a small portion of overall variance. Last, in order to correct for large pilot data, the original pilot sample size decreases the new suggested sample size. This formula approximation captures 96% of the variance in sample size scores and should allow a researcher to estimate based on their own data.

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert et al. [CITE, 2014]. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at OSF LINK [CITE]. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [CITE, Balota et al.]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here [https://elexicon.wustl.edu/, CITE]. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

>>>>

abstract: |

  Sample size planning for research studies often focuses on obtaining a significant result given a specified level of power, alpha, and proposed effect size. This planning generally requires prior knowledge of study design and a statistical analysis to calculate the proposed sample size. However, there may not be just one specific testable analysis from which to derive power [@silberzahn2018many] or even a hypothesis to test for the project (e.g., stimuli database creation). Newer power and sample size planning suggestions include Accuracy in Parameter Estimation [AIPE, @kelley2007, @maxwell2008] and simulation of proposed analyses [@chalmers2020]. These toolkits provide flexibility in traditional power analyses that focus on the if-this-then-that approach, yet, both AIPE and simulation require either a specific parameter (e.g., mean, effect size, etc.) or statistical test for planning sample size. In this tutorial, we explore how these latter two approaches can be combined to accommodate studies that may not have a specific hypothesis test or wish to account for the potential of a multiverse of analyses. Specifically, the examples focus on studies that implement multiple items and suggest that sample sizes can be planned to measure those items adequately and accurately, regardless of statistical test. Results show that pilot data can be used to determine a sample size that represents well-measured data, and multiple code vignettes will be provided for researchers to adapt and apply to their own measures.

@@

bibliography      : ["references.bib"]

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_docx
bibliography: references.bib
---

@@

Statistical power and power analyses are arguably one of the most important components to planning a research study [@cohen1990]. Yet, if reviews of transparency and openness in research publications are any clue, the social sciences have failed to fully implement power analyses as part of their common efforts [@hardwicke2020, @doi:10.1177/1745691620979806]. The replication "crisis" and credibility revolution have shown that many published studies in psychology are underpowered [@opensciencecollaboration2015, @vazire2018]. Pre-registration of a study involves outlining the study and hypotheses before data collection begins [@stewart2020, @d.chambers2014, @nosek2014] and then should summarily include a power analysis to determine the sample size necessary to detect the expected effect. Given the combined issues of publish-or-perish and that most non-significant results do not turn in published manuscripts, one may expect that power analysis would be especially critical for early career researchers [@rosenthal1979, @simmons2011]. Potentially, it is uninformative to publish studies that are underpowered [@halpern2002], but it can be difficult to know if a bad power analysis is better than no power analysis. A recent review of power analyses found in psychology journal articles indicates that researchers did not provide enough information to understand their power analyses and often chose effect sizes that were inappropriately justified [@beribisky2019].

One potential solution to the power analysis problem is the plethora of tools made available for researchers to simply power. G\*Power is one of the most popular free power software options [@faul2007, @erdfelder1996] that provides a simple point and click graphical user interface for power. Web-based tools have also sprung up for overall and statistical test specific sample size planning including powerandsamplesize.com and <https://designingexperiments.com> [@anderson2017]. Coding based packages, such as *pwr* [@champely2017], *faux* [@debruine2021], and *SimDesign* [@chalmers2020] can be used to examine power and sample size planning using *R*, usually with simulation. Researchers have to be careful using any toolkit, as errors can occur with the over-reliance on software [@nuijten2016]. When computing sample size estimates it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result [@williamson2000, @batterham2005].

Changes in publication practices and research design have also shown a new wrinkle to providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, multiple Many Analysts papers have shown that - given the same data and hypothesis - researchers can come up with multiple ways to analyze the data [@silberzahn2018many, @coretta]. Research projects often have multiple testable hypothesis, however, it is unclear which hypothesis or test the sample size planning should be performed on. Further, an entire set of common research publications may not even have a hypothesis to examine within their project, as they are simply providing a large, quality dataset for future reuse [i.e., stimuli database creation [@buchanan2019]. The increased ability to compute complex statistical analyses, such as multilevel modeling, has pushed researchers with repeated measures designs to abandon creating person-level averages just to be able to use a traditional ANOVA [@brysbaert2018]. These analyses have also made it clear that we should be careful to assume that all items in a research study have the "effect", as there is often variability in their impact on the outcomes of the study.

In this manuscript, we show a proposed method to account for variability in item effects, a potential lack of hypothesis test (or simply not a good way to estimate an effect size of interest), and/or an exploratory design with an unknown set of potential hypotheses and analyses choices. These methods are inspired by newer sample size planning methods including Accuracy in Parameter Estimation [AIPE, \@kelley2007, \@maxwell2008] and the ability to simulate proposed data for item estimates [@rousselet]. AIPE shifts the focus away from finding a *significant* *p*-value to finding a parameter that is "accurately measured". For example, a researcher may wish to detect a specific sized correlation in a study, *r* = .35. They could then use AIPE to estimate the sample size needed to find a "sufficiently narrow" confidence interval around that correlation. Sufficiently narrow is often defined by the researcher using a minimum parameter size of interest and confidence intervals. Therefore, they could decide that their 95% confidence interval should be approximately [.20, .50], and sufficiently narrow was defined as a width of .30 or .15 on each side. While confidence intervals are related to Null Hypothesis Significance Testing (i.e., 95% confidence intervals that do not include zero would indicate a significant difference from zero at $\alpha$ \< .05), AIPE procedures instead suggest a sample size that should obtain that width of a confidence interval, regardless if it includes zero.

```{r echo = F, eval = F}
library(MBESS)
ss.aipe.R2(Population.R2 = 0.35^2, # our correlation squared
           conf.level = 0.95, # confidence interval 
           width = 0.30^2, # the width squared
           p = 1) # one predictor X to Y
```

In this approach, a researcher could use sequential testing to estimate their parameter of interest after each participant to determine if they have achieved their expected width of the confidence interval around that parameter. In this procedure, one should consider setting a minimum sample size (i.e., based on known data collection ability), use the confidence interval width as a stopping rule (i.e., stop data collect when the CI is narrow), and use the estimated sample size from the AIPE calculations as a potential maximum sample size. By defining each of these components, a research could ensure a feasible minimum sample size, a way to cease data collection when goals have been met, and a stopping rule to ensure an actual end to data collection. Given pilot or previously collected data, we should be able to leverage the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for repeated measures studies with expected variabillity in parameter estimates for items.

@@

Using ideas from newer power methods, we suggest the following procedure to determine a sample size for each item:

@@

To explore if we could recover the new projected sample size from data a researcher would have, we used linear models to create a formula for researcher correction. First, the corrected projected sample size was predicted by the original projected sample size. Next, the standard deviation of the item standard deviations was added to the equation to recreate heterogeneity estimates. The scale of the data is embedded into the standard deviation of the items (*r* = `r round(cor(summary_long$scale, summary_long$sd_item), digits = 2)`), and therefore, this variable was not included separately. Last, we included the pilot sample size.

```{r regressions, include = F, echo = F}
user_model <- lm(new_sample ~ sample_size, data = summary_long)
user_print <- apa_print(user_model)
user_sum <- summary(user_model)

@@

The first model using pilot sample size to predict new sample size was significant, *F*(`r user_sum$fstatistic[2]`, `r user_sum$fstatistic[3]`) = `r round(user_sum$fstatistic[1], digits = 2)`, *p* \< .001, `r user_print$estimate$modelfit$r2`, capturing nearly 90% of the variance, `r user_print$estimate$sample_size`. The second model with item standard deviation was better than the first model *F*(`r change_table$df[2]`, `r format(change_table$res.df[2], big.mark = "")`) = `r round(change_table$statistic[2], digits = 2)`, *p* \< .001, `r user_print2$estimate$modelfit$r2`. The item standard deviation predictor was significant, `r user_print2$full_result$sd_item`. The addition of the original pilot sample size was also significant, *F*(`r change_table$df[3]`, `r format(change_table$res.df[3], big.mark = "")`) = `r round(change_table$statistic[3], digits = 2)`, *p* \< .001, `r user_print3$estimate$modelfit$r2`.

As shown in the final model Table \@ref(tab:table-correction), the new suggested sample size is proportional to the original suggested sample size (i.e., *b* \< 1), which reduces the sample size suggestion. As variability increases, the suggested sample size also increases to capture differences in heterogeneity shown above; however, this predictor is not significant in the final model, and only contributes a small portion of overall variance. Last, in order to correct for large pilot data, the original pilot sample size decreases the new suggested sample size. This formula approximation captures 96% of the variance in sample size scores and should allow a researcher to estimate based on their own data.

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert et al. [CITE, 2014]. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at OSF LINK [CITE]. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [CITE, Balota et al.]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here [<https://elexicon.wustl.edu/>, CITE]. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

```{=tex}
\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
```
::: {#refs custom-style="Bibliography"}
:::

```{=tex}
\endgroup
```
